Inference using context-dependent models of mutation

### 2021-10-12 writeups/context-dependence.tex
Finishing pass through section 3

The problem at hand is to infer the parameters of the model based on the final state of the model after it hsa evolved from some known initial state.
% If the sequence is long enough and not too much time has gone by
% this should be feasible -- there should be enough information in our observations to do this reliably.
% If the time is too large, the process may ``saturate'' --
% if there are many changes at most sites, then in at least most models,
% we will lose most information about the dynamics,
% retaining information only about the parameters that affect the stationary distribution.
% In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
% while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
% In the CpG model it is not immediately clear what information is retained.
How can we extract the information?
These are Markov processes, on the state space $\calS^n$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible for even moderately sized $n$.
We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, this ignores dependencies between neighboring blocks,
so it is not clear how good this approximation might be.
% Intuitively, we want to increase the size of the context
% -- and the likelihood thus obtained is asymptotically correct for $m$ large,
% but computationally, we are stuck with small values of $m$.

\begin{figure}
    \begin{center}
        \includegraphics{Tmer-dependency}
    \end{center}
    \caption{
        \textbf{(A)} A $(6,2,2)$ T-mer.
        Nonmutated sites are shown in lower case letters.
        The ``base'' of the T-mer is the lower central
        string of length $m = n - \ell - r$.
        \textbf{(B)}
        The three mutations have extended the influence of the circled \nA{}
        in the initial sequence (top)
        to the six positions colored red in the final sequence.
        The rate of each mutation depends on its' neighboring sites,
        i.e., with a window of width $w=1$.
        Red dotted arrows show the propagation of dependency
        between the circled \nA{}
        and the third site in the final sequence (bottom).
        \label{fig:Tmers}
    }
\end{figure}

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that something will be sufficient to compute the full likelihood
to an arbitrarily good approximation.
To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
the probability that an initial subsequence $x$ of length $n$, after time $t$,
is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
This central, smaller subsequence is called the ``base'' of the T-mer.
We refer to such patterns as ``T-mers'', as depicted in Figure~\ref{fig:Tmers}.
%EM Shall we define w here?
%PR No, I think that'll muddy the waters.
This quantity is found by marginalizing $p_m$,
summing over the possible inital and final substrings of $x$
(denoted $a$ and $b$):
\begin{align}
    p_{n,\ell,r}(t;x,y)
    &=
    \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) ,
    \quad
    \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
In Appendix~\ref{ss:approx_pf}, we show that this is so, namely that
the probability a given subsequence of length $n$ matching $x$
is later observed to have its center matching $y$
is close to $p_{n,\ell,m}(x,y)$, regardless of wider context.
Formally, for each $i$,
\begin{align} \label{eqn:window_approx}
    \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
    \approx
    p_{n,\ell,r}(x,y)
    .
\end{align}

%EM I feel like it needs emphasis that this is different than saying that we assume that the rates stay the same as in the initial sequence.
%PR I'm not sure what you mean here?

@@

What we actually want is the \emph{full likelihood} $\like(y|x)$,
the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
This can be computed as the product of the probability of each site in $y$
given $x$ and all sites in $y$ further to the left.
Defining $x_{<i}=(x_1, \ldots, x_{i-1})$ to be the subsequence to the left of $i$, this is:
\begin{align*}
    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\text{and}\; X=x \} .
\end{align*}
It turns out that
we can approximate the terms in this sum
by discarding approximately independent sites from the conditioning.
If $\ell$ is long enough that approximation \eqref{eqn:window_approx} holds, then:
\begin{align} \label{eqn:full_approx}
    \begin{split}
    \P\{ Y_i = y_i \given Y_{<i} \;\text{and}\; X \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        }
    \end{split} .
\end{align}
By rearranging the product,
this gives us an expression for the full likelihood as the ratio of the likelihoods of all $(n,\ell,\ell)$ T-mers in $(x,y)$
to that of of all $(n,\ell,\ell+1)$ T-mers.
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}
See Appendix~\ref{ss:approx_pf} for a more rigorous statement and proof,
and the next section for some intuition.

@@

The approximations in \eqref{eqn:window_approx} and \eqref{eqn:full_approx} hold
if the overhang, $\ell$, is long enough.
The intuition is straightforward, and provides a way to pick the overhang lengths $\ell$ and $r$;
a precise statement is given in Appendix \ref{ss:approx_pf}.
Suppose that each instantaneous change depends on positions at most $w$ sites away,
and refer to $w$ as the ``window width''.
Then, the mutation process at any particular site
depends on the initial sequence at that site and within distance $w$ on either side\ldots
\emph{and} the outcomes of any mutations that have occurred within that window.

>>>>

The problem at hand is to infer the parameters of the model based on the final state of the model after it has evolved from some known initial state.
% If the sequence is long enough and not too much time has gone by
% this should be feasible -- there should be enough information in our observations to do this reliably.
% If the time is too large, the process may ``saturate'' --
% if there are many changes at most sites, then in at least most models,
% we will lose most information about the dynamics,
% retaining information only about the parameters that affect the stationary distribution.
% In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
% while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
% In the CpG model it is not immediately clear what information is retained.
How can we extract the information?
These are Markov processes, on the state space $\calS^n$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible even for moderately sized $n$.
We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, this ignores dependencies between neighboring blocks,
so it is not clear how good this approximation might be.
% Intuitively, we want to increase the size of the context
% -- and the likelihood thus obtained is asymptotically correct for $m$ large,
% but computationally, we are stuck with small values of $m$.

%EM I suggest that we have something like (B) in the introduction.
%EM We need to make it clear that the window width of the underlying model, and the T-mer width, are two different things.
%EM Yes, this is in the last paragraph, but I think this point gets lost when we start with "a general context-dependent model."
\begin{figure}
    \begin{center}
        \includegraphics{Tmer-dependency}
    \end{center}
    \caption{
        \textbf{(A)} A $(6,2,2)$ T-mer.
        Nonmutated sites are shown in lower case letters.
        The ``base'' of the T-mer is the lower central
        string of length $m = n - \ell - r$.
        \textbf{(B)}
        The three mutations have extended the influence of the circled \nA{}
        in the initial sequence (top)
        to the six positions colored red in the final sequence.
        The rate of each mutation depends on its' neighboring sites,
        i.e., with a window of width $w=1$.
        Red dotted arrows show the propagation of dependency
        between the circled \nA{}
        and the third site in the final sequence (bottom).
        \label{fig:Tmers}
    }
\end{figure}

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that something will be sufficient to compute the full likelihood
to an arbitrarily good approximation.
To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
the probability that an initial subsequence $x$ of length $n$, after time $t$,
is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
This central, smaller subsequence is called the ``base'' of the T-mer.
We refer to such patterns as ``T-mers'', as depicted in Figure~\ref{fig:Tmers}.
%EM Shall we define w here?
%PR No, I think that'll muddy the waters.
This quantity is found by marginalizing $p_m$,
summing over the possible inital and final substrings of $x$
(denoted $a$ and $b$):
\begin{align}
    p_{n,\ell,r}(t;x,y)
    &=
    \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) ,
    \quad
    \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
In Appendix~\ref{ss:approx_pf}, we show that this is so, namely that
the probability a given subsequence of length $n$ matching $x$
is later observed to have its center matching $y$
is close to $p_{n,\ell,r}(x,y)$, regardless of wider context.
Formally, for each $i$,
\begin{align} \label{eqn:window_approx}
    \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
    \approx
    p_{n,\ell,r}(x,y)
    .
\end{align}

@@

What we actually want is the \emph{full likelihood} $\like(y|x)$,
the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
This can be computed as the product of the probability of each site in $y$
given $x$ and all sites in $y$ further to the left.
Defining $x_{<i}=(x_1, \ldots, x_{i-1})$ to be the subsequence to the left of $i$, this is:
\begin{align*}
    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \ \text{and}\; X=x \} .
\end{align*}
It turns out that
we can approximate the terms in this sum
by discarding approximately independent sites from the conditioning.
These approximately independent sites are exactly those that are too far away from a given site to impact its evolution.
Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
If $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds, then:
\begin{align} \label{eqn:full_approx}
    \begin{split}
    \P\{ Y_i = y_i \given Y_{<i} \ \text{and}\; X=x \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        }
    \end{split} .
\end{align}
By rearranging the product,
this gives us an approximation for the full likelihood as the ratio of the likelihoods of all $(2\ell+1,\ell,\ell)$ T-mers in $(x,y)$
to that of of all $(2\ell+1,\ell,\ell+1)$ T-mers.
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}
See Appendix~\ref{ss:approx_pf} for a more rigorous statement and proof,
and the next section for some intuition.
%EM I know that it's all written out in the appendix, but I think that having a little more development here would be nice.
%EM Can we re-instate the material just above, that's commented out?
%EM Also, what about the edge cases? When you say "all the t-mers", I think that includes t-mers where some of the "arms" are sticking off the sequence? Or are we making the sequence into a circle?

@@

The approximations in~\eqref{eqn:window_approx} and~\eqref{eqn:full_approx} hold
if the overhang, $\ell$, is long enough.
We present intuition about these formulas here, which provides a way to pick the overhang lengths $\ell$ and $r$;
a precise statement is given in Appendix~\ref{ss:approx_pf}.
Suppose that each instantaneous change depends on positions at most $w$ sites away,
and refer to $w$ as the ``window width''.
Then, the mutation process at any particular site
depends on the initial sequence at that site and within distance $w$ on either side\ldots
\emph{and} the outcomes of any mutations that have occurred within that window.
